---
title: "Reproducible Research: Peer Assessment 1"
author: Alex Pittendreigh
date: "17 October 2014"
output: 
  html_document:
    keep_md: true
---

## Introduction ##

It is now possible to collect a large amount of data about personal movement activity monitoring devices such as a [Fitbit](http://www.fitbit.com), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband) or [Jawbone Up](http://jawbone.com/up). However, the data generated by these devises tend to remain under-utilised because the data is hard to obtain and/or there is a lack of statistical methods and software for processing and interpreting this data.

This assessment will use a small sampling of data from one such device to answer questions regarding the activity performed over a two month period.

## Data ##

The device collects data at 5 minute intervals throughout the day and consists of two months of data from an anonymous individual gathered during the months of October and November in 2012. The original data set is located at [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) as at the 17th October 2014, however, this analysis will used the file provided in the Github repo provided by Dr Peng and is named **activity.zip**. The extracted raw data file from this archive is a comma separated value file called **activity.csv** and consists of 3 variables and 17568 observations.

The columns in this raw data set are:

1. **steps**: The number of steps taken in a 5 minute period with missing data indicated with NA  
2. **date**: Dates of the recorded activity in "YYYY-MM-DD" format  
3. **interval**: Identifier for the 5 minute interval in which the measurement was taken  

## Goal of Analysis ##

The broad questions that this analysis will answer are:

1. What is the mean total number of steps taken each day?
2. What is the average daily activity pattern?
3. Inputing missing values
4. Are there differences in activity patterns between weekdays and weekends?

This analysis will answer eaxh of these questions in order and provide a number of tables and plots,

## Analysis ##

This analysis assumes that the raw data file exists and that you have manually extracted it's contents into the current working directory.

### Loading and Preprocessing Data ###

Firstly lets load some needed libraries that will help manipulate the data

```{r}
library(plyr)
# No need to see the startup messages when loading dplyr
suppressPackageStartupMessages(library(dplyr))  
```

Then load the data and inspect the contents.

```{r}
activities <- tbl_df(read.csv('activity.csv'))

# Convert the 'date' column to a true date type
activities$date <- as.Date(activities$date)

activities
str(activities)
```

Looks to be pretty tidy data aside from the NA values which will be addressed shortly. We can proceed to perform the analysis for the first set of questions.

### Q1 - What is the mean total number of steps taken per day? ###

The first step is to summarise the data into a new dataframe giving the total number of steps per day. As the missing values can be hard to determine (i.e. they may be zero or may be plenty) we need to remove them from the summary data frame. After summarising the data, let's inspect the new dataframes values

```{r}
dailyStepsTally <- activities %>%
    filter(!is.na(steps)) %>%
    group_by(date) %>%
    summarise(totalSteps=sum(steps))

dailyStepsTally
```

Looks to be pretty valid information, so let's plot in out to gain a visual perspective of this data. The plotting system used is ggplot2 by Hadley Wickham.

```{r figure1, fig.width=10, fig.height=5}
library(ggplot2)

ggplot(subset(dailyStepsTally, !is.na(totalSteps)), aes(date, totalSteps)) +
    geom_bar(stat="identity", aes(fill=totalSteps)) +
    ggtitle("Total Daily Steps") +
    xlab("Date") +
    ylab("Number of Steps") +
    theme(axis.text.x=element_text(angle=90))
```

It would also be handy to know what †he mean and median values for the daily steps taken are:

```{r}
dailySteps.sum <- sum(dailyStepsTally$totalSteps)
dailySteps.mean <- format(mean(dailyStepsTally$totalSteps), digits = 9)
dailySteps.median <- median(dailyStepsTally$totalSteps)
```

| Sum | Mean | Median |
|:---:|:----:|:------:|
| `r dailySteps.sum` | `r dailySteps.mean` | `r dailySteps.median` |     

### Q2 - What is the average daily activity pattern?


